---
title: "PA1_template - updated"
author: "Jim Carson"
date: "March 26, 2017"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#John Hopkins Data Science Specialization
#Course 5: Reproducible Research
#Course Project 1

This assignment uses data from a personal activity monitoring device
Device collects data at 5 minute intervals through out the day
Two months of data from an anonymous individual collected Oct-Nov 2012

Dataset variables (17,568 observations)
  - steps: Number of steps taking in a 5-minute interval (missing values are coded as NA)
  - date: The date on which the measurement was taken in YYYY-MM-DD format
  - interval: Identifier for the 5-minute interval in which measurement was taken

```{r }
# Set working directory
setwd("~/R/John Hopkins Data Science Specialization/Course 5 - Reproducible Research/Data")

# Load graphics library
library("ggplot2", lib.loc="~/R/win-library/3.3")
library("lattice", lib.loc="~/R/win-library/3.3")

# Read data and review
Step_Data <- read.csv("activity.csv", header = TRUE)
str(Step_Data)
head(Step_Data)
class(Step_Data$date)

# Since 'date' field from factor to date
Step_Data$date <- as.Date(Step_Data$date)
```

#What is the mean total number of steps taken per day (ignore missing values)?
Calculate the total number of steps taken per day
Make a histogram of the total number of steps taken each day
Calculate and report the mean and median of the total number of steps taken per day
```{r }
Steps_per_day <- aggregate(steps ~ date, Step_Data, sum)
qplot(Steps_per_day$steps, geom="histogram", bins = 25,  fill=I("dark green"), col=I("dark green"), xlab="Number of Steps")
Mean_Steps_per_day <- round(mean(Steps_per_day$steps), 0)
Median_Steps_per_day <- round(median(Steps_per_day$steps), 0)
Mean_Steps_per_day
Median_Steps_per_day
```

The mean steps per day is `r Mean_Steps_per_day` and the median steps per day is `r Median_Steps_per_day`.

## What is the average daily activity pattern?
# Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)
# Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
```{r }
Steps_Interval <- aggregate(steps ~ interval, Step_Data, mean)
plot(Steps_Interval$interval,Steps_Interval$steps, type="l", xlab="Interval", ylab="Number of Steps")
Interval_Max <- Steps_Interval[which.max(Steps_Interval$steps),1]
Interval_Min <- Steps_Interval[which.min(Steps_Interval$steps),1]
Interval_Max
Interval_Min
```

The 5-minute interval with the maximum steps is `r Interval_Max`.

#Impute missing values
Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)
Devise a strategy for filling in all of the missing values in the dataset. 
Create a new dataset that is equal to the original dataset but with the missing data filled in.
Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. 
Do these values differ from the estimates from the first part of the assignment? 
What is the impact of imputing missing data on the estimates of the total daily number of steps?
```{r }
Missing_Count <- sum(!complete.cases(Step_Data))
Step_Data_Imputed <- transform(Step_Data, steps = ifelse(is.na(Step_Data$steps), Interval_Min, Step_Data$steps))
str(Step_Data_Imputed)

Steps_per_day_Imp <- aggregate(steps ~ date, Step_Data_Imputed, sum)
qplot(Steps_per_day_Imp$steps, geom="histogram", bins = 25,  fill=I("forest green"), col=I("forest green"), xlab="Number of Steps")
Mean_Steps_per_day_Imputed <- mean(Steps_per_day_Imp$steps)
Median_Steps_per_day_Imputed <- median(Steps_per_day_Imp$steps)

Mean_Diff <- round(Mean_Steps_per_day_Imputed - Mean_Steps_per_day, 0)
Median_Diff <- round(Median_Steps_per_day_Imputed - Median_Steps_per_day, 0)
Total_Diff <- round(sum(Steps_per_day_Imp$steps) - sum(Steps_per_day$steps), 0)
Mean_Diff
Median_Diff
Total_Diff
```

The difference in mean steps is `r Mean_Diff`, the difference in median steps is `r Median_Diff`, and the 
total difference is `r Total_Diff`.


#Are there differences in activity patterns between weekdays and weekends?
The weekdays() function may be of some help here. Use the dataset with the filled-in missing values for this part.
Create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.
Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). See the README file in the GitHub repository to see an example of what this plot should look like using simulated data.
```{r }
weekdays <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")
Step_Data_Imputed$Weekday = as.factor(ifelse(is.element(weekdays(as.Date(Step_Data_Imputed$date)),weekdays), "Weekday", "Weekend"))

Steps_Interval_Imputed <- aggregate(steps ~ interval + Weekday, Step_Data_Imputed, mean) 

xyplot(Steps_Interval_Imputed$steps ~ Steps_Interval_Imputed$interval|Steps_Interval_Imputed$Weekday, main="Average Steps per Day by Interval",xlab="Interval", ylab="Steps",layout=c(1,2), type="l")
```
